Community-Conditioned Language Models

### 2020-11-19 paper/paper.tex
partial refactor of tables
discussion and info gain

\documentclass[11pt,a4paper]{article}
\usepackage[hyperref]{acl2020}
\usepackage[T1]{fontenc}
\usepackage{times}
\usepackage{latexsym}
\usepackage{siunitx}
\usepackage{booktabs}
\usepackage{pgfplots}
\usepackage{todonotes}
\usepackage{enumitem}
\usepackage{cleveref}

\usepgfplotslibrary{groupplots}
\pgfplotsset{compat=1.16}

@@

For the LSTM, we trained the models on a simple left-to-right language modeling task
with cross entropy loss over the vocabulary. 
Because the transformer operates on all tokens in the sequence at once,
the language modeling was achieved by masking and incrementally un-masking input tokens
in a left-to-right fashion, following \citet{BERT}\jp{Or some other standard paper}.  
We used the AdamW \citep{Loshchilov2019} optimization algorithm, 
with an initial learning rate of \num{0.001}.\jp{Also for the transformer? No simulated annealing?}
We used a batch size of \num{256} and a maximum sequence length of \num{64} tokens, 
truncating longer messages (16.8\% of messages were longer than \num{64} tokens).
During training, a dropout rate of $0.1$ was applied between encoder layers 
and after each linear layer.

@@

@@

@@

@@

\begin{figure*}
\begin{tikzpicture}
  \confusiongroupplot{floats/lstm-3-1_comm_infer_confusion.csv}{floats/transformer-3-3_comm_infer_confusion.csv}
\end{tikzpicture}
\caption{LMCC confusion matrices for the best LSTM (left, $c=1$) and transformer (right, $c=3$) models.}
\label{fig:confusion}
\end{figure*}

@@

On the other hand, when we consider the \emph{information gain}\bn{define earlier}
of the CCLM over the corresponding (same encoder architecture)
un-conditioned

@@

We find that various communities exhibit widely varying levels of
linguistic complexity (\cref{hyp:varying-complexity}), with perplexity
per-word ranging from less than 53 to 104 for the best model
architecture.

@@

We find a weak correlation (TODO; r , p) between the perplexity of the
CCLM and the indiscernibility of the corresponding community
(\cref{hyp:rich-harder-to-identify}). Because we have relatively few,
and only large communities in our test set, it is difficult to
provide a conclusive answer, but this kind of correlation would suggest that
community-specific complexity can be partially attributed to using
more varied patterns. However, even the most complex communities
(perplexity per word above 100) have an indiscernibility of about 3
(out of a maximum of 46), so they retain fairly specific patterns:
they are not simply an even mixture of the kind of comments found in
other communities.

We find that the layer at which the community embedding is taken into
account in a CCLM influences its perplexity (\cref{hyp:layer-effect}),
albeit weakly. For LSTM models, the perplexity per word, averaged over
all communities, varies between 78.64 and 80.70. It is steadily
increasing when $l_c$ decreases.  For transformer models, it varies
between 89.38 and 122.56. Here perplexity per-word is not a monotonous
function of $l_c$: the maximum is reached at $l_c=2$. We did not
expect non-mononicity, and leave studying this effect to future work.

The representation of communities built by CCLM is exhibiting a clear
correlation with user-coocurrence patterns
(\cref{hyp:extra-linguistic-correlation}, TODO; r , p). A natural
question that may arise is how much this correlation can be explain by
user-specifc linguistic patterns. We plan to control for this variable
in future work, for example by building author-conditioned language
models and testing if community vector embeddings can be determined by
the set of the vector embeddings of its members.

- Comparison with topic modelling

The pattern of how much the perplexity falls across different communities
is similar across architectures, but with some differences.\jp{Either say something specific here or move the whole paragraph to future work.}
It could be interesting to further investigate these differences in future work,
since it could reveal differences in the kind of linguistic variation the different
model architectures capture.

>>>>

\documentclass[11pt,a4paper]{article}
\usepackage[hyperref]{acl2020}
\usepackage[T1]{fontenc}
\usepackage{times}
\usepackage{latexsym}
\usepackage{siunitx}
\usepackage{booktabs}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{todonotes}
\usepackage{enumitem}
\usepackage{array}
\usepackage{cleveref}

\usepgfplotslibrary{groupplots}
\pgfplotsset{compat=1.16}
\pgfplotstableset{
  column type=r,
  fixed zerofill,
  clear infinite,
  %dec sep align,
  %empty cells with={\ensuremath{-}},
}

@@

For the LSTM, we trained the models on a simple left-to-right language modeling task
with cross-entropy loss over the vocabulary. 
Because the transformer operates on all tokens in the sequence at once,
the language modeling was achieved by masking and incrementally un-masking input tokens
in a left-to-right fashion, following \citet{BERT}\jp{Or some other standard paper}.  
We used the AdamW \citep{Loshchilov2019} optimization algorithm, 
with an initial learning rate of \num{0.001}.\jp{Also for the transformer? No simulated annealing?}
We used a batch size of \num{256} and a maximum sequence length of \num{64} tokens, 
truncating longer messages (16.8\% of messages were longer than \num{64} tokens).
During training, a dropout rate of $0.1$ was applied between encoder layers 
and after each linear layer.

@@

%\pgfplotstabletypeset[
  %columns={best_epoch,lm_ppl,gain},
  %columns/model/.style={string type,column type=l},
  %columns/best_epoch/.style={column name={Epochs}, int detect},
  %columns/lm_ppl/.style={column name={Ppl}},
  %columns/gain/.style={column name={IG},precision=3,empty cells with={\ensuremath{-}}},
  %every nth row={5}{before row=\midrule},
  %]{floats/model.csv}

@@

We measure the perplexity of each example in the test set. 
For this, we use the negative exponential of the mean cross-entropy loss 
for the sequence.

@@

\subsection{Information gain of the CCLM}

We consider the average information gain of each CCLM over its baseline
un-conditioned counterpart with the same sequence encoder architecture.
For a given message, the information gain is defined as the ratio of the
perplexity of the baseline model to that of the CCLM:
\[
  IG_{CCLM}(m) = \frac{Ppl_{LM}(m)}{Ppl_{CCLM}(m)}
\]

\begin{table*}
  \centering
  \input{floats/info_gain.tex}
  \caption{Mean information gain of the CCLM over the baseline model, for messages in the test set.}
  \label{tab:model-info-gain}
\end{table*}

While the absolute performance of the LSTM models is better,
we find the the transformer CCLMs have higher information gain 
\cref{tab:model-info-gain}.

@@

\begin{figure*}
\begin{tikzpicture}
  \confusiongroupplot{lstm-3-1}{transformer-3-3}
\end{tikzpicture}
\caption{LMCC confusion matrices for the best LSTM (left, $c=1$) and transformer (right, $c=3$) models.}
\label{fig:confusion}
\end{figure*}

@@

On the other hand, community indiscernibility is correlated with
the average information gain of the CCLM over the baseline model,
for messages in its test set \cref{fig:comm-pca}.
Communities for which the language model bennifits most from 
the conditioned model are generally easier for the LMCC to classify.

@@

We find that various communities exhibit widely varying levels of
linguistic complexity \cref{hyp:varying-complexity}, with perplexity
per-word ranging from less than 53 to 104 for the best model
architecture.

@@

We find a weak correlation \cref{tab:cclm_lmcc_ppl} between the perplexity of the \bn{I'm not sure if we can make this claim or not with the Ppl-Ppl correlations. However, given that *all* of them have positive correlation, even though the p-values are somewhat higher than .1 maybe we can?}
CCLM and the indiscernibility of the corresponding community
(\cref{hyp:rich-harder-to-identify}). Because we have relatively few,
and only large communities in our test set, it is difficult to
provide a conclusive answer, but this kind of correlation would suggest that
community-specific complexity can be partially attributed to using
more varied patterns. However, even the most complex communities
(perplexity per word above 100) have an indiscernibility of about 3
(out of a maximum of 46), so they retain fairly specific patterns:
they are not simply an even mixture of the kind of comments found in
other communities.

We find that the layer at which the community embedding is taken into
account in a CCLM influences its perplexity (\cref{hyp:layer-effect}),
albeit weakly. For LSTM models, the perplexity per word, averaged over
all communities, varies between 76.80 and 78.64. \bn{80.70 is the baseline model} 
The perplexity per word is slightly higher for $l_c=0$, and has very
little difference between the other models, with $l_c=1$ being the best.
For transformer models, it varies
between 89.38 and 122.56. Here perplexity per-word is clearly not a monotonous
function of $l_c$: the maximum is reached at $l_c=2$. We did not
expect non-mononicity, and leave studying this effect to future work.

The pattern of information gain by community is similar across architectures.
Communities that bennifit most from the conditioned model behave that way for 
both the LSTM and transformer.
However, there are some differences.
For example, /r/relationships has an information gain of \num{1.018} for
the best CCLM LSTM but only \num{0.998} for the best transformer model. \bn{todo: community-wise CCLM info gain}
Why was the LSTM able to take advantage of the community information for
this community, where the transformer wasn't?
It would be interesting to investigate these differences in future work,
since it could reveal differences in the kind of linguistic variation the different
model architectures capture.

The representation of communities built by CCLM exhibits a clear
correlation with user-coocurrence patterns
(\cref{hyp:extra-linguistic-correlation}). A natural
question that may arise is how much this correlation can be explain by
user-specifc linguistic patterns. We plan to control for this variable
in future work, for example by building author-conditioned language
models and testing if community vector embeddings can be determined by
the set of the vector embeddings of its members.
